Local Spectral Clustering of Density Upper Level Sets

### 2019-05-10 arxiv/arxiv_local_density_clustering_proofs.tex
updated mixing time proof and section, edited other sections

\newcommand{\diam}{\mathrm{diam}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\defeq}{\overset{\mathrm{def}}{=}}
\newcommand{\vol}{\mathrm{vol}}
\newcommand{\abs}[1]{\left \lvert #1 \right \rvert}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Reals}{\mathbb{R}}
\newcommand{\Rd}{\Reals^d}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\1}{\mathbf{1}}
\newcommand{\var}{\mathrm{Var}}
\newcommand{\Err}{\mathrm{Err}}
\newcommand{\Log}{\mathrm{Log}}
\newcommand{\TV}{\mathrm{TV}}
\newcommand{\dist}{\mathrm{dist}}
\newcommand{\Id}{\mathrm{Id}}

%%% Graph terms
\newcommand{\cut}{\mathrm{cut}}
\newcommand{\degminpr}{\deg_{\min}'}
\newcommand{\degminwt}{\widetilde{\deg}_{\min}}
\newcommand{\degmaxwt}{\widetilde{\deg}_{\max}}
\newcommand{\degmax}{\deg_{\max}}
\newcommand{\piminwt}{\widetilde{\pibf}_{\min}}
\newcommand{\piminpr}{\pibf_{\min}'}
\newcommand{\degmin}{\deg_{\min}}

@@

%%% Operators
\DeclareMathOperator*{\argmin}{arg\,min}

@@

%%% Tilde notation for quantities over the expansion set 
\newcommand{\wn}{\widetilde{n}}
\newcommand{\wX}{\widetilde{\Xbf}}
\newcommand{\wx}{\widetilde{x}}
\newcommand{\wz}{\widetilde{z}}
\newcommand{\wbz}{\widetilde{\bf{z}}}
\newcommand{\wu}{\widetilde{u}}
\newcommand{\wPbb}{\widetilde{\Pbb}}
\newcommand{\wf}{\widetilde{f}}
\newcommand{\wDbf}{\widetilde{\Dbf}}

\newtheoremstyle{aldenthm}
{6pt} % Space above
{6pt} % Space below
{\itshape} % Body font
{} % Indent amount
{\bfseries} % Theorem head font
{.} % Punctuation after theorem head
{.5em} % Space after theorem head
{} % Theorem head spec (can be left empty, meaning `normal')

\theoremstyle{aldenthm}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}

@@

We will need to carefully control the volume of expansion sets using the estimate in Lemma \ref{lem: expansion_volume}; Lemma \ref{lem: Taylor_series} serves this purpose.
\begin{lemma}
	\label{lem: Taylor_series}
	For any $0 \leq x \leq 1/2d$,
	\begin{align*}
	(1 + x)^d & \leq 1 + 2dx \\
	(1 - x)^d & \geq 1 - 2dx.
	\end{align*}
\end{lemma}
\begin{proof}
	We take the binomial expansion of $(1 + x)^d$:
	\begin{align*}
	(1 + x)^d & = \sum_{k = 0}^{d} {d \choose k} x^k \\
	& = 1 + dx + dx\left(\sum_{k = 2}^d \frac{{d \choose k} x^{k - 1}}{d}\right)\\
	& \leq 1 + dx + dx\left(\sum_{k = 2}^d \frac{{d \choose k}}{(2d)^{k-1}d}\right) \tag{since $x \leq \frac{1}{2d}$} \\
	& \leq 1 + dx + dx\left(\sum_{k = 2}^d \frac{1}{2^{k - 1}}\right) \leq 1 + 2dx.
	\end{align*}

@@

\begin{lemma}
	\label{lem: expected_number_boundary_points}
	Under the conditions of Theorem \ref{thm: conductance_upper_bound}, and for any $0 < r \leq \sigma/2d$,
	\begin{equation*}
	\Pbb(\Csigr) \leq \frac{2dr}{\sigma} \left(\lambda_{\sigma} - c_0\frac{r^{\gamma}}{\gamma + 1}\right) \nu(\Csig)
	\end{equation*}	
\end{lemma}
\begin{proof}
	We partition $\Csigr$ into slices based on distance from $\Csig$ as follows: for $k \in \N$,
	\begin{equation*}
	\mathcal{T}_{i,k} = \set{x \in \Csigr: t_{i,k} < \frac{\dist(x, \Csig)}{r} \leq t_{i+1,k}}, ~~ \Csigr = \bigcup_{i = 0}^{k-1} \mathcal{T}_{i,k}
	\end{equation*}
	where $t_i = i/k$ for $i = 0, \ldots, k - 1$. As a result, for any $k \in \mathbb{N}$,
	\begin{equation}
	\label{eqn: partition_ub}
	\Pbb(\Csigr) = \int_{\Csigr} f(x) dx = \sum_{i = 0}^{k-1} \int_{\mathcal{T}_{i,k}} f(x) dx \leq \sum_{i = 0}^{k-1} \nu(\mathcal{T}_{i,k}) \max_{x \in \mathcal{T}_{i,k}} f(x).
	\end{equation}

@@

Lemma \ref{lem: local_conductance} will be necessary to lower bound the expected volume of $\Csig[\Xbf]$ in $G_{n,r}$. Define the \emph{local conductance} $\ell_{\nu,r}(u)$ to be
\begin{equation*}
\ell_{\nu,r}(u) = \frac{\nu\left(\Csig \cap B(u,r)\right)}{\nu\left(B(u,r)\right)}.
\end{equation*}
\begin{lemma}
	\label{lem: local_conductance}
	Let $u \in \Csig$. Then, for any $0 < r \leq \frac{\sigma}{2\sqrt{d}}$,
	\begin{equation*}
	\ell_{\nu,r}(u) \geq \frac{6}{25}. 
	\end{equation*}
\end{lemma}
\begin{proof}
	Since $u \in \Csig$ there exists $x \in \Cset$ such that $u \in B(x, \sigma)$, and as $B(x,\sigma) \subseteq \Csig$,
	\begin{equation*}
	\nu\bigl(B(u, r) \cap B(x, \sigma)\bigr) \leq \nu\bigl(B(u, r) \cap \Csig \bigr).
	\end{equation*}
	Without loss of generality, let $\norm{u - x} = \sigma$; it is not hard to see that if $\norm{u - x} < \sigma$, the volume of the overlap will only grow. Then, since $\norm{u  - x} = \sigma$, $B(u, r) \cap B(x, \sigma)$ contains a spherical cap of radius $r$ and height
	\begin{equation*}
	h = r - (r)^2/2\sigma = r \left( 1 - \frac{r}{2 \sigma} \right)
	\end{equation*}	
	which by Lemma \ref{lem: volume_of_spherical_cap} has volume
	\begin{equation*}
	\nu_{cap} = \frac{1}{2} \nu_d r^d I_{1 - \alpha}\left( \frac{d + 1}{2}  ,\frac{1}{2}\right)
	\end{equation*}
	with $\alpha = 1 - \frac{2rh - h^2}{r^2} = \frac{r^2}{4 \sigma^2} \leq \frac{1}{8d}$.

@@

\begin{lemma}
	\label{lem: expected_density_cut}
	Under the setup and conditions of Theorem \ref{thm: conductance_upper_bound}, and for any $0 < r \leq \sigma/2d$,
	\begin{equation*}
	p_K \leq \frac{4 d \nu_d r^{d+1} \lambda}{\sigma} \left(\lambda_{\sigma} - c_0\frac{r^{\gamma}}{\gamma + 1}\right) \nu(\Csig)
	\end{equation*}
\end{lemma}
\begin{proof}
	We can write $\cut_{n,r}$ as a double sum,
	\begin{equation}
	\label{eqn: density_cut_expansion}
	\cut_{n,r} = \sum_{i = 1}^{n} \sum_{j \neq i} \1(x_i \not\in \Csig) \1(x_j \in \Csig) \1(\norm{x_i - x_j} \leq r)
	\end{equation}
	and by linearity of expectation, we obtain
	\begin{equation*}
	p_K = \frac{\mu_K}{{n \choose 2}} = 2 \cdot \Pbb(x_i \not\in \Csig, x_j \in \Csig, \norm{x_i - x_j} \leq r). \tag{for each $i,j$, $i \neq j$}
	\end{equation*}
	Writing this with respect to the density function $f$, we have
	\begin{align*}
	p_K & = 2 \int_{\Rd \setminus \Csig} f(x) \Pbb\bigl(B(x,r) \cap \Csig\bigr) dx \\
	& = 2 \int_{\Csigr} f(x) \Pbb\bigl(B(x,r) \cap \Csig\bigr) dx \\
	& \leq 2 \nu_d r^d \lambda  \int_{\Csigr} f(x) dx = 2 \nu_d r^d \lambda \Pbb(\Csigr).
	\end{align*}
	where the inequality follows from \ref{asmp: cluster_separation}, which implies $f(x) \leq \lambda$ for $x \in \Csig \setminus \Cset$. Then, upper bounding the integral using Lemma \ref{lem: expected_density_cut} gives the final result.
\end{proof}

\begin{lemma}
	\label{lem: expected_density_volume}
	Under the setup and conditions of Theorem \ref{thm: conductance_upper_bound}, and for any $0 < r \leq \sigma/2d$,
	\begin{equation*}
	p_V \geq \frac{12}{25} \lambda_{\sigma}^2 \nu_d r^d \nu(\Csig)
	\end{equation*}
\end{lemma}
\begin{proof}
	The proof will proceed similarly to Lemma \ref{lem: expected_density_cut}. We begin by writing $\vol_{n,r}$ as the sum of indicator functions,
	\begin{equation}
	\label{eqn: volume_expansion}
	\vol_{n,r} = \sum_{i = 1}^{n} \sum_{j \neq i} \1(x_i \in \Csig) \1(x_j \in B(x_i, r))
	\end{equation}
	and by linearity of expectation we obtain
	\begin{equation*}
	p_V = \frac{\mu_V}{{n \choose 2}} = 2 \cdot \Pbb(x_i \in \Csig, x_j \in B(x_i,r)). \tag{for any $i,j$, $i \neq j$. }
	\end{equation*}
	Writing this with respect to the density function $f$, we have
	\begin{align*}
	p_V & = 2 \int_{\Csig} f(x) \Pbb(B(x,r)) dx \\
	& \geq 2 \int_{\Csig} f(x) \Pbb(B(x,r) \cap \Csig) dx
	\end{align*}
	whence the claim then follows by Lemma \ref{lem: local_conductance}.
\end{proof}

@@

For $N \in \mathbb{N}$ and a set $V$ of $N$ vertices, take $G = (V,E)$ to be an undirected and unweighted graph, with associated adjacency matrix $\Abf$, random walk matrix $\Wbf$, and stationary distribution $\pibf = (\pi_u)_{u \in V}$ where $\pi_v = \frac{\Dbf_{vv}}{\vol(V; G)}$. For $v \in V$, 
\begin{equation}
\label{eqn: random_walk}
q_{vu}^{(m)} = e_v\Wbf^m e_u, ~~ \qbf_{v}^{(m)} = \left(q_{vu}^{(m)}\right)_{u \in V}, ~~ \qbf_v = (\qbf^{(1)}_{v\cdot},\qbf^{(2)}_{v\cdot}, \ldots), 
\end{equation}
denote respectively the $m$-step transition probability, distribution, and sequence distributions of the random walk over $G$ originating at $v$. Letting $\qbf = (\qbf_v)_{v \in V}$, the relative pointwise mixing time is thus
\begin{equation*}
\tau_{\infty}(\qbf; G) := \tau_{\infty}(\qbf; G) = \min\set{m: \forall u,v \in V, \frac{\abs{q_{vu}^{(m)} - \pibf_u}}{\pibf_u} \leq 1/4} 
\end{equation*}
where we include the extra dependency on $\qbf$ as we will need to consider mixing of various random walks.

Two key quantities relate the mixing time to the expansion of subsets $S$ of $V$. The \emph{local spread} is defined to be
\begin{equation*}
s(G) := \frac{9D_{\min}}{10}\pi_{\min} 
\end{equation*}
for $D_{\min} := \min_{v \in V} \Dbf_{vv}$ and $\pi_{\min} := D_{\min} / \vol(V; G)$.

%\inf \set{t: \forall S \subset V ~\text{with}~ \pi(S) = t, \beta(S) \leq 1/10}
%\end{equation*}
where $\beta(S) := \inf_{v \in S} \qbf_{v}^{(1)}(S^c)$, and by convention we let $\pbf(S) = \sum_{u \in S} p_u$ for any distribution vector $\pbf = (p_u)_{u \in V}$ over $V$. We collect some necessary facts about the local spread in Lemma \ref{lem: local_spread_G}.
\begin{lemma}
	\label{lem: local_spread_G}
	\begin{itemize}
		\item If $\pibf(S) \leq s(G)$, then for every $u \in S$, $\qbf_u^{(1)}(S^c) \geq 1/10$.
		\item For any $v, u \in V$, and $m \in N$ greater than $0$, $q_{vu}^{(m)}/ \pi_{\min} \leq 1/s(G)$.
	\end{itemize}

@@

\begin{proof}
	If $t = \pibf(S) \leq \frac{9 D_{\min}}{10} \pi_{\min}$, divide both sides by $\pi_{\min}$ to obtain
	\begin{equation*}
	\abs{S} \leq \frac{9 D_{\min}}{10}
	\end{equation*}
	which implies $\qbf_{v}^{(1)}(S^c) \geq 1/10$ for all $v \in S$. This implies the first statement.

	The second statement follows from the fact $q_{vu}^{(m)} \leq 1/D_{\min}$ for any $m$.
\end{proof}

The local spread facilitates conversion between $\tau_{\infty}(\qbf_v; G)$ and the more easily manageable \emph{total variation} mixing time, given by
\begin{equation*}
\tau_1(\rhobf; G) = \min\biggl\{m: \forall v \in V, \norm{\rhobf_v - \pibf}_{TV} \leq 1/4 \biggr\}
\end{equation*}
where 
\begin{equation}
\label{eqn: uniform_random_walk}
\rhobf_{v}^{(m)} = \frac{1}{m}\sum_{k = 1}^{m+1} \qbf_{v}^{m}, ~~ \rhobf_v = \left( \rhobf_{v}^{(1)}, \rhobf_{v}^{(2)}, \rhobf_{v}^{(3)} \ldots \right), ~~ \rhobf = \left( \rhobf_v \right)_{v \in V}
\end{equation}
and $\norm{\pbf - \pibf}_{TV} = \sum_{v \in V}\abs{p_v - \pi_v}$ is the total variation norm between distributions $\pbf$ and $\pibf$. 
\begin{lemma}
	\label{lem: tv_mixing_to_pointwise_mixing1}
	For $\qbf$ as in \eqref{eqn: random_walk} and $\rhobf$ as in \eqref{eqn: uniform_random_walk},
	\begin{equation*}
	\tau_{\infty}(\qbf; G) \leq 2752 \tau_1(\rhobf; G) \log \left(4 \max\left\{1, \frac{1}{s(G)}\right\}\right)
	\end{equation*}
\end{lemma}
\begin{proof}
	Masking dependence on the starting vertex $v$ for the moment, let
	\begin{equation*}
	\Delta_u^{(m)} = q_{vu}^{(m)} - \pi_u, ~~ \delta_u^{(m)} = \frac{\Delta_u^{(m)}}{\pi_u}
	\end{equation*}
	and $\Deltabf^{(m)} = (\Delta_u^{(m)})_{u \in V}$, $\deltabf^{(m)} = (\delta_u^{(m)})_{u \in V}$. For a vector $\Deltabf = (\Delta_u)_{u \in V}$, the $L^{p}(\pibf)$ norm is given by
	\begin{equation*}
	\norm{\Deltabf}_{L^p(\pibf)} = \left(\sum_{u \in V} \left(\Delta_u\right)^{p} \pi_u \right)^{1/p}
	\end{equation*}
	To go between the $L^{\infty}(\pibf)$ and $L^{1}(\pibf)$ norms, we have
	\begin{align*}
	\norm{\deltabf^{(2m)}}_{L^{\infty}(\pi)} & \overset{(i)}{\leq} \norm{\deltabf^{(m)}}^2_{L^{2}(\pi)} \\
	& = \norm{(\deltabf^{(m)})^2}_{L^{1}(\pi)} \\
	& \overset{(ii)}{\leq}  \norm{(\deltabf^{(m)})}_{L^{1}(\pi)} \norm{(\deltabf^{(m)})}_{L^{\infty}(\pi)}
	\end{align*}
	where $(i)$ is a result of \cite{benjamini2003} and $(ii)$ follows from Holder's inequality. Now, we upper bound the second factor on the right hand side by observing
	\begin{align*}
	\norm{(\deltabf^{(m)})}_{L^{\infty}(\pi)} & \leq \max\left\{1, \max_{u \in V} \frac{q_{vu}^{(m)}}{\pi_u} \right\} \\
	& \overset{(iii)}{\leq} \max\left\{1, \frac{1}{s(G)}\right\}
	\end{align*}
	where $(iii)$ follows from Lemma \ref{lem: local_spread_G}.

>>>>

\newcommand{\diam}{\mathrm{diam}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\seq}[1]{\left\{#1\right\}_{n \in \mathbb{N}}}
\newcommand{\defeq}{\overset{\mathrm{def}}{=}}
\newcommand{\vol}{\mathrm{vol}}
\newcommand{\abs}[1]{\left \lvert #1 \right \rvert}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Reals}{\mathbb{R}}
\newcommand{\Rd}{\Reals^d}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\1}{\mathbf{1}}
\newcommand{\var}{\mathrm{Var}}
\newcommand{\Err}{\mathrm{Err}}
\newcommand{\Log}{\mathrm{Log}}
\newcommand{\TV}{\mathrm{TV}}
\newcommand{\dist}{\mathrm{dist}}
\newcommand{\Id}{\mathrm{Id}}

%%% Graph terms
\newcommand{\cut}{\mathrm{cut}}
\newcommand{\degminpr}{\deg_{\min}'}
\newcommand{\degminwt}{\widetilde{\deg}_{\min}}
\newcommand{\degmaxwt}{\widetilde{\deg}_{\max}}
\newcommand{\degmax}{\deg_{\max}}
\newcommand{\piminwt}{\widetilde{\pi}_{\min}}
\newcommand{\piminpr}{\pibf_{\min}'}
\newcommand{\degmin}{\deg_{\min}}

@@

%%% Operators
\DeclareMathOperator*{\argmin}{arg\,min}
\newcommand{\dx}{\,dx}
\newcommand{\dy}{\,dy}
\newcommand{\dt}{\,dt}

@@

%%% Tilde notation for quantities over the expansion set 
\newcommand{\wn}{\widetilde{n}}
\newcommand{\wX}{\widetilde{\Xbf}}
\newcommand{\wx}{\widetilde{x}}
\newcommand{\wz}{\widetilde{z}}
\newcommand{\wbz}{\widetilde{\bf{z}}}
\newcommand{\wu}{\widetilde{u}}
\newcommand{\wPbb}{\widetilde{\Pbb}}
\newcommand{\wf}{\widetilde{f}}
\newcommand{\wDbf}{\widetilde{\Dbf}}
\newcommand{\piwt}{\widetilde{\pi}}

\newtheoremstyle{aldenthm}
{6pt} % Space above
{6pt} % Space below
{\itshape} % Bo\dy font
{} % Indent amount
{\bfseries} % Theorem head font
{.} % Punctuation after theorem head
{.5em} % Space after theorem head
{} % Theorem head spec (can be left empty, meaning `normal')

\theoremstyle{aldenthm}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}

@@

We will need to carefully control the volume of expansion sets using the estimate in Lemma \ref{lem: expansion_volume}; Lemma \ref{lem: Taylor_series} serves this purpose.
\begin{lemma}
	\label{lem: Taylor_series}
	For any $0 \leq x \leq 1/2d$,
	\begin{align*}
	(1 + x)^d & \leq 1 + 2\dx \\
	(1 - x)^d & \geq 1 - 2\dx.
	\end{align*}
\end{lemma}
\begin{proof}
	We take the binomial expansion of $(1 + x)^d$:
	\begin{align*}
	(1 + x)^d & = \sum_{k = 0}^{d} {d \choose k} x^k \\
	& = 1 + \dx + \dx\left(\sum_{k = 2}^d \frac{{d \choose k} x^{k - 1}}{d}\right)\\
	& \leq 1 + \dx + \dx\left(\sum_{k = 2}^d \frac{{d \choose k}}{(2d)^{k-1}d}\right) \tag{since $x \leq \frac{1}{2d}$} \\
	& \leq 1 + \dx + \dx\left(\sum_{k = 2}^d \frac{1}{2^{k - 1}}\right) \leq 1 + 2\dx.
	\end{align*}

@@

\begin{lemma}
	\label{lem: expected_number_boundary_points}
	Under the conditions of Theorem \ref{thm: conductance_upper_bound}, and for any $0 < r \leq \sigma/2d$,
	\begin{equation*}
	\Pbb(\Csigr) \leq \frac{2dr}{\sigma} \left(\lambda_{\sigma} - c_0\frac{r^{\gamma}}{\gamma + 1}\right) \nu(\Csig)
	\end{equation*}	
\end{lemma}
\begin{proof}
	We partition $\Csigr$ into slices based on distance from $\Csig$ as follows: for $k \in \N$,
	\begin{equation*}
	\mathcal{T}_{i,k} = \set{x \in \Csigr: t_{i,k} < \frac{\dist(x, \Csig)}{r} \leq t_{i+1,k}}, ~~ \Csigr = \bigcup_{i = 0}^{k-1} \mathcal{T}_{i,k}
	\end{equation*}
	where $t_i = i/k$ for $i = 0, \ldots, k - 1$. As a result, for any $k \in \mathbb{N}$,
	\begin{equation}
	\label{eqn: partition_ub}
	\Pbb(\Csigr) = \int_{\Csigr} f(x) \dx = \sum_{i = 0}^{k-1} \int_{\mathcal{T}_{i,k}} f(x) \dx \leq \sum_{i = 0}^{k-1} \nu(\mathcal{T}_{i,k}) \max_{x \in \mathcal{T}_{i,k}} f(x).
	\end{equation}

@@

Lemma \ref{lem: local_conductance} will be necessary to lower bound the expected volume of $\Csig[\Xbf]$ in $G_{n,r}$. Define the \emph{uniform local conductance} $\ell_{\nu,r}(u)$ to be
\begin{equation*}
\ell_{\nu,r}(u) = \nu\bigl(\Csig \cap B(u,r)\bigr)
\end{equation*}
\begin{lemma}
	\label{lem: local_conductance}
	Let $u \in \Csig$. Then, for any $0 < r \leq \frac{\sigma}{2\sqrt{d}}$,
	\begin{equation*}
	\ell_{\nu,r}(u) \geq \frac{6}{25} \nu_d r^d.
	\end{equation*}
\end{lemma}
\begin{proof}
	Since $u \in \Csig$ there exists $x \in \Cset$ such that $u \in B(x, \sigma)$, and as $B(x,\sigma) \subseteq \Csig$,
	\begin{equation*}
	\nu\bigl(B(u, r) \cap B(x, \sigma)\bigr) \leq \nu\bigl(B(u, r) \cap \Csig \bigr)
	\end{equation*}
	Without loss of generality, let $\norm{u - x} = \sigma$; it is not hard to see that if $\norm{u - x} < \sigma$, the volume of the overlap will only grow. Then, since $\norm{u  - x} = \sigma$, $B(u, r) \cap B(x, \sigma)$ contains a spherical cap of radius $r$ and height
	\begin{equation*}
	h = r - (r)^2/2\sigma = r \left( 1 - \frac{r}{2 \sigma} \right)
	\end{equation*}	
	which by Lemma \ref{lem: volume_of_spherical_cap} has volume
	\begin{equation*}
	\nu_{cap} = \frac{1}{2} \nu_d r^d I_{1 - \alpha}\left( \frac{d + 1}{2}  ,\frac{1}{2}\right)
	\end{equation*}
	with $\alpha = 1 - \frac{2rh - h^2}{r^2} = \frac{r^2}{4 \sigma^2} \leq \frac{1}{8d}$.

@@

\begin{lemma}
	\label{lem: expected_density_cut}
	Under the setup and conditions of Theorem \ref{thm: conductance_upper_bound}, and for any $0 < r \leq \sigma/2d$,
	\begin{equation*}
	p_K \leq \frac{4 d \nu_d r^{d+1} \lambda}{\sigma} \left(\lambda_{\sigma} - c_0\frac{r^{\gamma}}{\gamma + 1}\right) \nu(\Csig)
	\end{equation*}
\end{lemma}
\begin{proof}
	We can write $\cut_{n,r}$ as a double sum,
	\begin{equation}
	\label{eqn: density_cut_expansion}
	\cut_{n,r} = \sum_{i = 1}^{n} \sum_{j \neq i} \1(x_i \not\in \Csig) \1(x_j \in \Csig) \1(\norm{x_i - x_j} \leq r)
	\end{equation}
	and by linearity of expectation, we obtain
	\begin{equation*}
	p_K = \frac{\mu_K}{{n \choose 2}} = 2 \cdot \Pbb(x_i \not\in \Csig, x_j \in \Csig, \norm{x_i - x_j} \leq r). \tag{for each $i,j$, $i \neq j$}
	\end{equation*}
	Writing this with respect to the density function $f$, we have
	\begin{align*}
	p_K & = 2 \int_{\Rd \setminus \Csig} f(x) \Pbb\bigl(B(x,r) \cap \Csig\bigr) \dx \\
	& = 2 \int_{\Csigr} f(x) \Pbb\bigl(B(x,r) \cap \Csig\bigr) \dx \\
	& \leq 2 \nu_d r^d \lambda  \int_{\Csigr} f(x) \dx = 2 \nu_d r^d \lambda \Pbb(\Csigr).
	\end{align*}
	where the inequality follows from \ref{asmp: cluster_separation}, which implies $f(x) \leq \lambda$ for $x \in \Csig \setminus \Cset$. Then, upper bounding the integral using Lemma \ref{lem: expected_density_cut} gives the final result.
\end{proof}

\begin{lemma}
	\label{lem: expected_density_volume}
	Under the setup and conditions of Theorem \ref{thm: conductance_upper_bound}, and for any $0 < r \leq \sigma/2d$,
	\begin{equation*}
	p_V \geq \frac{12}{25} \lambda_{\sigma}^2 \nu_d r^d \nu(\Csig)
	\end{equation*}
\end{lemma}
\begin{proof}
	The proof will proceed similarly to Lemma \ref{lem: expected_density_cut}. We begin by writing $\vol_{n,r}$ as the sum of indicator functions,
	\begin{equation}
	\label{eqn: volume_expansion}
	\vol_{n,r} = \sum_{i = 1}^{n} \sum_{j \neq i} \1(x_i \in \Csig) \1(x_j \in B(x_i, r))
	\end{equation}
	and by linearity of expectation we obtain
	\begin{equation*}
	p_V = \frac{\mu_V}{{n \choose 2}} = 2 \cdot \Pbb(x_i \in \Csig, x_j \in B(x_i,r)). \tag{for any $i,j$, $i \neq j$. }
	\end{equation*}
	Writing this with respect to the density function $f$, we have
	\begin{align*}
	p_V & = 2 \int_{\Csig} f(x) \Pbb(B(x,r)) \dx \\
	& \geq 2 \int_{\Csig} f(x) \Pbb(B(x,r) \cap \Csig) \dx
	\end{align*}
	whence the claim then follows by Lemma \ref{lem: local_conductance}.
\end{proof}

@@

For $N \in \mathbb{N}$ and a collection $V = \set{v_1,\ldots,v_N}$ of $N$ vertices, take $G = (V,E)$ to be an undirected and unweighted graph, with associated adjacency matrix $\Abf$, degree matrix $\Dbf$, and random walk matrix $\Wbf$. We consider the stationary distribution $\pi$ to be a mapping from $V$ to $[0,1]$, with $\pi(u) := \frac{\Dbf_{uu}}{\vol(V; G)}$ for $u \in V$. For $v \in V$, let $q: V \times V \to [0,1]$ represent the $m$-step transition probability of the random walk over $G$ originating from $v$, $q^{(m)}(v,u) = e_v\Wbf^m e_u$. The relative pointwise mixing time over $G$ is then
\begin{equation*}
\tau_{\infty}(G) = \min\set{m: \forall u,v \in V, \frac{\abs{q^{(m)}(v,u) - \pi(u)}}{\pi(u)} \leq 1/4}. 
\end{equation*}

Two key quantities relate the mixing time to the expansion of subsets $S$ of $V$. The \emph{local spread} is defined as
\begin{equation*}
s(G) := \frac{9}{10} \cdot \min_{v \in V} \set{\deg(v; G)} \cdot \min_{v \in V} \set{\pi(v)}
\end{equation*}
For $f: V \to \Reals$, and $S \subseteq V$, write $f(S) := \sum_{u \in S} f(u)$. We collect some necessary facts about the local spread in Lemma \ref{lem: local_spread_G}.
\begin{lemma}
	~

	\label{lem: local_spread_G}
	\begin{itemize}
		\item If $\pi(S) \leq s(G)$, then for every $u \in S$, $q^{(1)}(u,S^c) \geq \frac{1}{10}$.
		\item For any $v, u \in V$, and $m \in \mathbb{N}, m \geq 0$,
		\begin{equation*}
		q^{(m)}(v,u) \leq \frac{\min_{v \in V} \set{\pi(v)}}{s(G)}.
		\end{equation*}
	\end{itemize}

@@

\begin{proof}
	If $\pi(S) \leq s(G)$, divide both sides by $\min_{v \in V} \set{\pi(v)}$ to obtain $\abs{S} \leq 9\min_{v \in V} \set{\deg(v; G)}/10$,
	which implies $q^{(1)}(v,S^c) \geq 1/10$ for all $v \in S$. This implies the first statement.

	The second statement follows from the fact $q^{(m)}(v,u) \leq 1/\min_{v \in V} \set{\deg(v; G)}$ for any $m$.
\end{proof}

The local spread facilitates conversion between $\tau_{\infty}(G)$ and the more easily manageable \emph{total variation} mixing time, given by
\begin{equation*}
\tau_1(G) = \min\biggl\{m: \forall v \in V, \norm{\rho^{(m)}(v,\cdot) - \pi}_{TV} \leq 1/4 \biggr\}
\end{equation*}
where $\rho^{(m)}(u,v) = \sum_{l = 1}^{m} p^{(l)}(u,v)$ 
and $\norm{\rho(v,\cdot) - \pi}_{TV} = \sum_{u \in V}\abs{\rho(u,v) - \pi(v)}$ is the total variation norm between distributions $\rho(v,\cdot)$ and $\pi$. 
\begin{lemma}
	\label{lem: tv_mixing_to_pointwise_mixing1}
	\begin{equation*}
	\tau_{\infty}(G) \leq 2752 \cdot \tau_1(G) \log \left(4 \max\left\{1, \frac{1}{s(G)}\right\}\right)
	\end{equation*}
\end{lemma}
\begin{proof}
	Masking dependence on the starting vertex $v$ for the moment, let
	\begin{equation*}
	\Delta^{(m)}(u) = q^{(m)}(v,u) - \pi(u), ~~ \delta^{(m)}(u) = \frac{\Delta^{(m)}(u)}{\pi(u)}
	\end{equation*}
	and denote $\Deltabf^{(m)} = \bigl(\Delta^{(m)}(u)\bigr)_{u \in V}$, $\deltabf^{(m)} = \bigl(\delta^{(m)}(u)\bigr)_{u \in V}$. For a vector $\Deltabf = (\Delta_u)_{u \in V}$, the $L^{p}(\pi)$ norm is given by
	\begin{equation*}
	\norm{\Deltabf}_{L^p(\pi)} = \left(\sum_{u \in V} \left(\Delta_u\right)^{p} \pi(u) \right)^{1/p}
	\end{equation*}
	To go between the $L^{\infty}(\pi)$ and $L^{1}(\pi)$ norms, we capitalize on a clever argument developed in \cite{morris2005}. For any $v,u \in V$, by the time-reversibility of the graph random walk,
	\begin{align}
	\abs{\frac{q^{(2m)}(v,u) - \pi(u)}{\pi(u)}} & = \abs{\sum_{y \in V} \pi(y) \left( \frac{q^{(m)}(v,y) - \pi(y)}{\pi(y)}\right)\left(\frac{q^{(m)}(u,y) - \pi(y)}{\pi(y)} \right)} \nonumber \\
	& \leq \norm{(\deltabf^{(m)})}_{L^{1}(\pi)} \norm{(\deltabf^{(m)})}_{L^{\infty}(\pi)}, \label{eqn: infty_norm_to_one_norm}
	\end{align}
	with the second line following by Holder's inequality.

	Now, we upper bound the second factor on the right hand side by observing
	\begin{align}
	\norm{(\deltabf^{(m)})}_{L^{\infty}(\pi)} & \leq \max\left\{1, \max_{u \in V} \frac{q^{(m)}(v,u)}{\pi(u)} \right\} \nonumber \\
	& \overset{(i)}{\leq} \max\left\{1, \frac{1}{s(G)}\right\} \label{eqn: ub_infinity_norm}
	\end{align}
	where $(i)$ follows from Lemma \ref{lem: local_spread_G}.

	Set $m_0 = 1376 \tau_1(G) $, which we will show implies $\norm{q^{(m)}(v,\cdot) - \pi}_{TV} \leq \frac{1}{4}$. Now, we leverage the following well-known fact (see pg. 1 of \cite{montenegro2002}): for any $\epsilon > 0$, if $m \geq m_0 \cdot \log(1/\epsilon)$ then for any $v \in V$,
	\begin{equation}
	\label{eqn: rapid_mixing}
	\norm{q^{(m)}(v,\cdot) - \pi}_{TV} \leq \epsilon.
	\end{equation}
	But $\norm{q^{(m)}(v,\cdot) - \pi}_{TV}$ is exactly $\norm{\deltabf^{(m)}}_{L^{1}(\pi)}$. Letting
	\begin{equation*}
	m_1 = m_0 \cdot \log \left(4 \max\left\{1, \frac{1}{s(G)}\right\} \right)
	\end{equation*} 
	by \eqref{eqn: infty_norm_to_one_norm}, \eqref{eqn: ub_infinity_norm}, and \eqref{eqn: rapid_mixing}, $\norm{(\deltabf^{(m)})}_{L^{\infty}(\pi)} \leq 1/4$ for all $m \geq 2 m_1$.  Recall that
	\begin{equation*}
	\norm{(\deltabf^{(m)})}_{L^{\infty}(\pi)} = \sup_{u \in V}\left\{ \frac{\abs{q^{(m)}(v,u) - \pi(u)}}{\pi(u)} \right\}.
	\end{equation*}
	and since none of the above depended on a specific choice of $v$, the supremum can be taken over all starting vertices $v$ as well. Thus for all $m \geq 2m_1$, for any $v,u \in V$, $\abs{q^{m}(v,u) - \pi(u)} \leq \frac{\pi(u)}{4}$.

	Finally, it is known (pg.15 of \citep{montenegro2002}) that
	\begin{equation*}
	\norm{\rho^{(m)}(v,\cdot) - \pi}_{TV} \leq \frac{1}{4} \Longrightarrow  \norm{q^{(1376m)}(v,\cdot) - \pi}_{TV} \leq \frac{1}{4} 
	\end{equation*}
	and so $m_0 = 1376 \tau_1(G)$ implies $\norm{q^{(m)}(v,\cdot) - \pi}_{TV} \leq \frac{1}{4}$, completing the proof.
\end{proof}
